---
title: "Text analytics & topic modelling on song lyrics"
author: ''
description: Text analytics & topic modelling in more than 380,000 song lyrics, of
  various music genres in the past 50 years
slug: text-analytics-topic-modelling-songs-lyrics
tags:
- Text analytics
- Tidytext
- LDA
- Topic modelling
categories: R
---

The main objective is to **develop clusters of decades/albums by the song lyrics** 
and the steps are the following:

* Data preparation (cleansing, transform etc.)
* Exploratory analysis  
* Topic modelling

Various R libraries were used, but it is mainly based on [#tidytext](https://twitter.com/hashtag/tidytext?src=hash) 
and [#tidyverse](https://twitter.com/hashtag/tidyverse?src=hash) environment.


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Libraries
library(readr)
library(tidyverse)
library(stringr)
library(tidytext)
library(lubridate)
library(wordcloud)
library(topicmodels)
library(tm)
library(stopwords)
library(quanteda)
library(ggthemes)
library(cld3)
```

Data preparation was a very important part of the analysis. 
- detecting english by using the **cld3** library, the
origin of the songs was detected & added to the dataset. The language detection wasn't perfect, as it 
misclassified a few songs. Almost all songs were correctly detected except for a couple of NA results and the "lala -lyric" song already identified in previous analysis as completely nonesense was assigned "ja" so I decided to remove it. The NA results were actually english, thus I kept them for the analysis. 

```{r}
setwd("C:/Users/lavin/Desktop/Bob Dylan Project/Bob Workflow")
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Insert data
bob <- read.csv("BobDylanDataset.csv", header = TRUE, sep= ";", stringsAsFactors = FALSE)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# DATA CLEANSING
## Filter data, handle missing values
# fix contractions
fix.contractions <- function(doc) {
  doc <- gsub("won't", "will not", doc)
  doc <- gsub("can't", "can not", doc)
  doc <- gsub("n't", " not", doc)
  doc <- gsub("'ll", " will", doc)
  doc <- gsub("'re", " are", doc)
  doc <- gsub("'ve", " have", doc)
  doc <- gsub("'m", " am", doc)
  doc <- gsub("'d", " would", doc)
  # 's could be 'is' or could be possessive: it has no expansion
  doc <- gsub("'s", "", doc)
  return(doc)
}
# fix (expand) contractions
bob$lyric <- sapply(bob$lyric, fix.contractions)
# function to remove special characters, if any
removeSpecialChars <- function(x) gsub("[^a-zA-Z0-9 ]", " ", x)
# remove special characters
bob$lyric <- sapply(bob$lyric, removeSpecialChars)
# convert everything to lower case
bob$lyric <- sapply(bob$lyric, tolower)

#create buckets and group the years into decades
bob <- bob %>%
  mutate(decade = 
           ifelse(bob$year %in% 1962:1969, "1960s", 
                  ifelse(bob$year %in% 1970:1979, "1970s", 
                         ifelse(bob$year %in% 1980:1989, "1980s", 
                                ifelse(bob$year %in% 1990:1999, "1990s", 
                                       ifelse(bob$year %in% 2000:2009, "2000s",
                                              ifelse(bob$year %in% 2010:2017, "2010s",
                                                     "NA")))))))
#REMOVE INSTRUMENTAL SONGS 
bob <- bob[!(bob$lyric=="instrumental"),]

# Detect the language of the song
library(cld3)
bob$lang <- detect_language(bob$lyric)
# Filter the songs that 
bob_lan <- 
  bob %>%
  filter(lang == "en" | lang== NA )
  
bob$characters <- str_count(bob$lyric)
```


```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
## _text cleaning
# Create a vector with stopwords
stopwords <- c(stopwords())
  
# Clean text
bob$lyric <- tolower(bob$lyric)
bob$lyric <- removePunctuation(bob$lyric)
bob$lyric <- removeNumbers(bob$lyric)
bob$lyric <- stripWhitespace(bob$lyric)
bob$lyric <- removeWords(bob$lyric, stopwords)
bob$lyric <- stemDocument(bob$lyric)
# Save processed data for future use
saveRDS(bob, file = "cleaned_bob.RDS")
```


```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
## _Insert processed dataset
songs <- readRDS(file = "cleaned_bob.RDS")
```

# MAIN ANALYSIS

Below there is a statistical table and a frequency plot to indicate the 
differences between the decades. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# library(DT)
# 
# knit_print.data.frame = function(x, ...) {
#   knit_print(DT::datatable(x), ...)
# }
songs_data <- 
songs %>% 
  group_by(decade) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(`Proportion(%)` = round((n/sum(n))*100, 2)) %>% 
  arrange(-n) %>% 
  rename(Decade = decade,
         `Total songs` = n) 
knitr::kable(songs_data, caption = "Songs per decade")
#knit_print.data.frame()
```



```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
songs %>% 
  group_by(decade) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(Freq = n/sum(n)) %>% 
  arrange(-Freq) %>% 
  ggplot() +
  geom_col(aes(reorder(decade, -Freq), Freq), fill = "steelblue", alpha = 0.7) +
  labs(y = "Number of songs", x = "Decade", 
       title = "Proportion of songs per Decade", 
       subtitle = "Bob Dylan Song Collection")+
  theme_fivethirtyeight() +
  scale_y_continuous(labels = scales::percent_format()) 
  
```
From the table is clear how the 60s and 70s account for more than 50% of all Bob Dylan Songs. 


Below, differences between albums for clarity sake 

```{r echo=TRUE}
album_data <- 
songs %>% 
  group_by(album) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(`Proportion(%)` = round((n/sum(n))*100, 2)) %>% 
  arrange(-n) %>% 
  rename(Album = album,
         `Total songs` = n) 
knitr::kable(album_data, caption = "Songs per Album")
```


```{r echo=TRUE}
songs %>% 
  group_by(album) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(Freq = n/sum(n)) %>% 
  arrange(-Freq) %>% 
  ggplot() +
  geom_col(aes(reorder(album, -Freq), Freq), fill = "steelblue", alpha = 0.7) +
  labs(y = "Number of songs", x = "Album", 
       title = "Proportion of songs per Album", 
       subtitle = "Bob Dylan Song COllection")+
  theme_fivethirtyeight() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  scale_y_continuous(labels = scales::percent_format()) 
```
Albums such as Triplicate, The Basement Tapes and Self Portrait account for the majority of total songs proportion. Their tracklist is simply longer. 



Now let's try to figure out which music genre uses more lyrics.


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
## number of characters per song
songs %>% 
  ggplot() +
  geom_boxplot(aes(decade, characters), fill = "steelblue", alpha = 0.7) +
  labs(y = "Length of song lyrics (in characters)", x = "Music Genre", 
       title = "Length (in characters) of song lyrics per decade", 
       subtitle = "From 1970 to 2016")+
  theme_fivethirtyeight() +
  ylim(0, 10000)

songs %>% 
  group_by(decade) %>% 
  summarise(characters = round(mean(characters, na.rm = TRUE), 0)) %>% 
  ggplot(aes(reorder(decade, -characters), characters)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  labs(y = "Length of song lyrics (in characters)", x = "Music Genre", 
       title = "Average lyrics characters per decade", 
       subtitle = "From 1970 to 2016")+
  theme_fivethirtyeight()
```



```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
## number of characters per song
songs %>% 
  ggplot() +
  geom_boxplot(aes(album, characters), fill = "steelblue", alpha = 0.7) +
  labs(y = "Length of song lyrics (in characters)", x = "album", 
       title = "Length (in characters) of song lyrics per Album", 
       subtitle = "From 1962 to 2017")+
  theme_fivethirtyeight() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylim(0, 10000)

songs %>% 
  group_by(album) %>% 
  summarise(characters = round(mean(characters, na.rm = TRUE), 0)) %>% 
  ggplot(aes(reorder(album, -characters), characters)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  labs(y = "Length of song lyrics (in characters)", x = "Album", 
       title = "Average lyrics characters per Album", 
       subtitle = "From 1970 to 2016")+
  theme_fivethirtyeight() +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))

```


# WORDCLOUD

It is interesting to see which are the most used words in each music genre. 
Below there are word clouds for the top 5 music genres. 
Word clouds (also known as text clouds or tag clouds) work in a simple way: 
the more a specific word appears in a source of textual data, the bigger 
and bolder it appears in the word cloud.  
Below there are word clouds for decades.


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
words <- 
  songs %>%
  unnest_tokens(word, lyric) %>% 
  group_by(decade, word) %>% 
  count() %>% 
  arrange(-n) %>% 
  group_by(decade) %>% 
  top_n(n = 100, wt = n)
# Select top 5 genres
decades <- c("1960s", "1970s", "1980s", "1990s", "2000s", "2010s")
for(i in 1:length(decades)){
  temp <- filter(words, decade == decades[i])
  
  # Create a word cloud
  par(bg="grey30")
  wordcloud(words = temp$word, freq = temp$n, col=terrain.colors(length(temp$word), alpha=0.9), random.order=FALSE, rot.per=0.3 )
  title(main =  decades[i] , font.main = 1, col.main = "cornsilk3", cex.main = 1.2)
}
  
```



# TOPIC MODELLING

Want to do is to classify documents into topics or themes. 
Among other things this would help you figure out if there’s anything interest 
while also directing you to the relevant subsets of the corpus. For small 
collections, one could do this by simply going through each document but this 
is clearly unfeasible for corpuses containing thousands of documents.

**Topic modeling** deals with the problem of automatically
classifying sets of documents into themes. The algorithm chosen is 
**Latent Dirichlet Allocation or LDA**, which essentially is a technique that 
facilitates the automatic discovery of themes in a collection of documents.

The basic assumption behind LDA is that each of the documents in a collection 
consist of a mixture of collection-wide topics. However, in reality we observe 
only documents and words, not topics – the latter are part of the hidden (or latent) 
structure of documents. The aim is to infer the latent topic structure given the
words and document. LDA does this by recreating the documents in the corpus by
adjusting the relative importance of topics in documents and words in topics 
iteratively.

In our case an LDA model with two topics was developed. After computing the topic
probabilities for all songs, we can see if this unsupervised learning, distinguish 
or reveal associations between music genres (regarding their lyrics).  
The box-plot below, reveals the probabilities of each music genre song to belong 
in each of the three topics.


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# _Build Model ###############################################################
  
# split into words
by_word <- 
  songs %>% 
  unnest_tokens(word, lyric)
# find document-word counts
word_counts <- 
  by_word %>%
  count(ï..song, word, sort = TRUE) %>%
  ungroup()
# Create document term matrix
songs_dtm <- word_counts %>%
  cast_dtm(ï..song, word, n)
songs_lda <- LDA(songs_dtm, k = 3, control = list(seed = 1234))
# Save for future use
save(songs_lda, file = "Bob_lda_3.RDA")
``` 


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load the model
load(file = "Bob_lda_3.RDA")
# _Calculate Tables ##########################################################
library(tidytext)
  
ap_topics <- tidy(songs_lda, matrix = "gamma")
  
topics_probs <- right_join(ap_topics, songs[, c("ï..song", "decade")], by = c("document" = "ï..song"))
 
topics_probs %>%
  mutate(decade = reorder(decade, gamma * topic)) %>%
  ggplot(aes(factor(topic), gamma, colour = factor(topic))) +
  geom_boxplot(alpha = 0.7) +
  labs(y = "Probability", x = "Topic", 
       title = "Topic probabilities per decade", 
       subtitle = "") +
  theme_fivethirtyeight() +
  labs(col="Topics",
       y = "Probabilities") +
  facet_wrap(~ decade)
```

